$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-12-25 qdpm_paper.tex
Worked on text

This paper is organized as follows. In \secref{sec:Background} we provide relevant background for the QDM and related models. In \secref{sec:QDPM} we present the details of the QPDM. \secref{sec:numerics} provides the results of a Monte Carlo study of the ground state of the QDPM at its exactly soluble point. Finally, \secref{sec:discussion} discusses potential future work.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.9\linewidth]{example_local_constraints.pdf}
    \caption{Examples of GQDMs with different local constraints. Clockwise from the upper left panel; fixed dimer number: $n_d=1$, $n_d=2$;  fixed dimer parity: odd, even.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires only $(Q_v\mod 2)$ to vanish.

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -\sum_\nu t_\nu \hat{T}_{\nu} + \sum_\nu v_\nu \hat{V}_{\nu} \label{eq:HGQDM}
\end{equation}
where $\hat{T}_\nu$ are off-diagonal kinetic energy operators, $\hat{V}_\nu$ are diagonal the potential energy operators, $t_\nu$ and $v_\nu$ parametrize the strength of these terms, and each $\nu$ and represents a set of  neighboring links. We will limit our discussion to $t_\nu >0$. $\hat{T}_\nu$ is constructed to generate resonances between two distinct allowed dimerazations of $\nu$, $c_\nu$ and its compliment $\bar{c}_\nu$, and is written as:
\begin{equation}
\hat{T}_{\nu} = \eket{c_\nu}\ebra{\bar{c}_\nu} + h.c.;
\end{equation}
in the simplest case, $\hat{T_\nu}$ involves the minimum non-trivial dimer dynamics that preserve the constraint. Similarly, the potential energy term provides an energy cost for states with these local configurations:
\begin{equation}
\hat{V}_{\nu} = \eket{c_\nu}\ebra{c_\nu} +  \eket{\bar{c}_\nu}\ebra{\bar{c}_\nu}.
\end{equation}
Therefore we may define a Hamiltonian for a GQDM using \Eqref{eq:HGQDM} and defining a set of dimer resonances $\{(c_\nu,\bar{c}_\nu)\}$ and corresponding potential and kinetic energy strenghts $\{(t_\nu,v_\nu)\}$:
\begin{equation}
\HGQDM \biggl( \bigl\{\left(c_\nu,\bar{c}_\nu\right)\},\{\left(t_\nu,v_\nu\right)\bigr\}\biggr).
\end{equation}
 In general the $T_\nu$'s don't commute with each other or with the $V_\nu$'s, and thus the ground state of $H_{\rm{GQDM}}$ is not explicitly known.

However, at the special points in the $\{(t_\nu,v_\nu)\}$ parameter space, known as the Rokhsar-Kivelson (RK) points,the exact ground state is known when \Eqref{eq:HGQDM} can be written as the sum of projection operators
\begin{equation}
H_{\rm{RK}} = \sum_\nu \hat{h}_\nu \label{eq:HRK},
\end{equation}
where all $h_\nu$ are projectors with eigenvalues $\{0,1\}$; a zero energy ground state of \Eqref{eq:HRK} is simultaneously annihilated by all $\hat{h}_\nu$. Such a zero energy state, known as the RK state, may be formed from an equal superposition of all dimerizations that are connected by $\{T_\nu\}$:
\begin{equation}
\eket{\Phi_{\RK}^{\Omega}} = \sum_{C \in \Omega} \eket{C}
\end{equation} 
where $\Omega$ is a subset of the Hilbert space connected by $\{T_\nu\}$. Thus $\{ \ket{\Phi_{\RK}^\Omega} \}$ are the exact, degenerate ground states of $H_{\rm{GQDM}}$ with a degeneracy labelled by $\Omega$ at the RK point.

@@

If a defect is introduced which violates the local constraint at a vertex, this can be viewed as an ``electric charge'' living on the vertex, within the context of the original GQDM. One may determine if such electric charges are confined by introducing a pair of such defects and computing their spatial correlation function; if the free energy of defects a large distances remains finite, electric charges are deconfined in the GQDM ground state.

The gauge symmetry allows for the construction of a gauge invariant magnetic string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell \label{eq:phi}
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv - \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) \tilde{n}_\ell \label{eq:phiBackground}
\end{equation}
where $n_\ell$ is the number operator for a static background dimer configuration which is chosen to fix the gauge, and the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma+ \tilde{\Phi}_\Gamma$. (Alternately can think of the source of $\tilde{\Phi}_\Gamma$ to be static charges of charge $+c_0$ ($-c_0$) sitting at the verticies of the $A$($B$) sub-lattice.).

We define the string operator is defined as
\begin{equation}
S_\gamma \left( \alpha \right) \equiv e^{i \alpha \left( \Phi_\gamma - \tilde{\Phi}_\gamma \right)} \label{eq:Sgamma}
\end{equation}
where $\alpha$ such that $S_\gamma$ commutes with $\HGQDM$ everywhere except at the ends of $\gamma$. Therefore, for closed loops $\Gamma$, $[ S_\Gamma,\HGQDM]=0$. For open strings, $S_\gamma$ does not commute with the Hamiltonian and can be viewed as creating a pair of ``magnetic charges'' at the ends of the string: correspondingly $S_\gamma \ket{\Phi_{\RK}}$ can used as a trial wavefunction for an excited state with a pair of magnetic vortices. If a state has topological order with deconfined electric charges, these magnetic vortices should be gapped excitations, and correspondingly the magnetic string operator should decay exponentially~\cite{Read1989a,Senthil2000,Senthil2001e}. The condensation of magnetic vortices can drive a phase transition from a topologically ordered phase to a conventionally order phase~\cite{Jalabert1991,Ralko2007,Huh2011}. Thus, the magnetic string correlation can be used as test for deconfinement of electric charges.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tpb]
    \centering
    \includegraphics[width=1.0\columnwidth]{mag_loops_on_orntd_lat.pdf}
    \caption{$\Phi_i$: a nontrivial string winding around the system in one of the two possible
        directions. $\Phi_2$: an open
    magnetic string with two different charged (colored) visons at the end points. $\Phi_3$ a trivial closed
    string. The colors red and blue distinguish the two different sublattices that orient the links. \note{Owen: Add a dimer/pentamer-ization to the figure. Also, add a monomer defect in the magnetic loop, to demonstrate the fractional statistics of defects.}}
    \label{fig:orientedSL}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The two unique fixed-number constraints on the square lattice are lead to QDM in the odd sector and the quantum full-packed loop model (QFPLM) in the even sector. For configurations that satisfy these a number constraints, $Q_v$ strictly vanishes, and thus all states in the Hilbert space are invariant under $G_v$ for $0 \leq \alpha_v \leq 2 \pi$ in \Eqref{eq:Gv}; this represents a local $U(1)$ gauge symmetry.

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes $\bplqt$ of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have taken $t_\bplqt = t$ \& $v_\bplqt =v$ $\forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0
\end{equation}
and therefore we can define two conserved topological winding numbers $\Phi_{x}$ \& $\Phi_{y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On an lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_{x/y}$ are integers in the range
\begin{equation}
-L_{x/y}/2 \leq  \Phi_{x/y} \leq L_{x/y}/2,
\end{equation}
thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

At the Rokhsar-Kivelson point $t=v$, we may write $\HQDM$ in the form of \Eqref{eq:HRK} by defining projection operators  $\hat{h}_\nu \equiv \hat{P}^\dagger_{\nu} \hat{P}_{\nu}$, with
\begin{equation}
\hat{P}^\dagger_\nu \equiv \biggl( \eket{ c_\nu} - \eket{\bar{c}_\nu} \biggr).
\end{equation} 
Since the plaquette flip term in \Eqref{eq:HQDM} is ergodic in each of topological sector $(\Phi_x,\Phi_y)$ on the square lattice (\textcolor{red}{is this known for the FPLM???})~\cite{????}, each topological sector corresponds to a unique RK ground state where $\Omega$ is labeled by $(\Phi_x,\Phi_y)$. Thus the equal superposition RK points of the QDM and QFPLM have an \emph{extensively} degenerate ground states.

Defects in these models correspond to states with $Q_v \neq 0$; a table of the charges associated with defects in the QDM and QFPLM is listed in \tabref{tab:defects}.

The QDM and QFPLM on the square lattice have isolated liquid ground states at the RK point surrounded by symmetry broken phases~\cite{Rokhsar1988,Leung1996,Syljuasen2005,Syljuasen2006,Syljuasen2004,Moessner2008}. The RK point of these models have been shown to have power-law decaying dimer-dimer and monomer-monomer correlation functions and gapless excitations~\cite{Fisher1961,Kasteleyn1961,Fisher1966a,Fisher1963,Kasteleyn1963a,Sutherland1988b,Sutherland1988a,Kohmoto1988a,Liang1988,Sutherland1988,Kohmoto1988,Levitov1990,Leung1996,Henley1997,Henley1997,Henley2004a,Pollmann2006}. Thus, such fixed number constrained models on bipartite lattices are expected to have an isolated unstable gapless critical quantum liquid points with logarithmically confined defect and an extensive topological degeneracy. These features of fixed number constrained models are consistent with the effective U(1) gauge theory description of the RK point~\cite{Fradkin2004,Fradkin1990,Moessner2001}; in 2+1D pure U(1) gauge theories are always confining and thus the lack of an extended deconfined topologically ordered phase near the RK point~\cite{PolyakovBook}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \begin{tabular}{| l | c | c | c | c | c |}
  	\hline
    %type & $n_v$ & $Q_v^{\rm{QDM}}$ & $Q_v^{\rm{FPLM}}$  & $Q_v^{Z_3}$ & \\
    type & $n_v$ & $Q_v^{\rm{QDM}}$ & $Q_v^{\rm{FPLM}}$  & $Q_v^{QDPM}$ & \\
	\hline
    monomer & $0$ & -1 & -2 &  -1 & $\svrtmon$  \\
	\hline
    dimer & $1$ & 0 & -1 &   0 & $\svrtdmr$ \\
	\hline
    trimer & $2$ & 1 & 0 &   1 & $\svrttrmr$ $\svrttrmrtwo$ \\
	\hline
    tetramer & $3$ & 2 & 1 &   -1 & $\svrtquat$ \\
	\hline
    pentamer & $4$ & 3 & 2 &  0 & $\svrtpent$  \\
	\hline
  \end{tabular}
  \caption{Defects where $n_v$ dimers touch a vertex and their associated $U(1)$ charges on the A sub-lattice in both the QDM, QFPLM, and QDPM. Defects on the B sub-lattice have opposite charge.}
  \label{tab:defects}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We now consider relaxing the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) (and the closely related even and odd sector Toric Codes). In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{IGT}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{IGT}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include rotational symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

The increased dynamics of $\HIGT$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HIGT$. Consequently, to define the string and loop operators which commute with $\HIGT$ (for closed strings) we fix $\alpha = \pi$ in \Eqref{eq:Sgamma}:
\begin{equation}
S_\gamma^{\rm{IGT}} \equiv S_\gamma \left( \pi \right), \quad W_{X,Y}^{\rm{IGT}} \equiv W_{X,Y} \left( \pi \right).
\end{equation}
Thus the $W_{X,Y}^{\rm{IGT}}$ have two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as $\Omega: (\pm 1,\pm1)$; consequently the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

Defects in the Ising Gauge theories have $Q_v^{Z_2} = 1$ and we identify these defects as $Z_2$ electric charges on the corresponding vertices.

The ground states of $\HIGT$ are fourfold degenerate on the torus and have exponentially decaying dimer-dimer and monomer-monomer correlations (\textcolor{red}{Is this known for both sectors}) and a finite energy gap; such features demonstrated the deconfined topologically ordered nature of the ground state of the IGTs.

>>>>

This paper is organized as follows. In \secref{sec:Background} we provide relevant background for the QDM and related models. In \secref{sec:QDPM} we present the details of the QPDM. \secref{sec:numerics} provides the results of a Monte Carlo study of the ground state of the QDPM at its exactly soluble point. Finally, \secref{sec:discussion} discusses potential future studies of the QDPM.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.9\linewidth]{example_local_constraints.pdf}
    \caption{Examples of GQDMs with different local constraints. Clockwise from the upper left panel; fixed dimer number: $n_d=1$, $n_d=2$;  fixed dimer parity: odd, even.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link is either unoccupied or occupied by a single dimer. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires $(Q_v\mod 2)$ to vanish.

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -\sum_\nu t_\nu  {T}_{\nu} + \sum_\nu v_\nu  {V}_{\nu} \label{eq:HGQDM}
\end{equation}
where ${T}_\nu$ are off-diagonal kinetic energy operators, ${V}_\nu$ are diagonal potential energy operators, $t_\nu$ and $v_\nu$ parametrize the strength of these terms, and each $\nu$ and represents a set of  neighboring links. We will limit our discussion to $t_\nu >0$. $ t{T}_\nu$ is constructed to generate resonances between two distinct allowed dimerazations of $\nu$, $c_\nu$ and its compliment $\bar{c}_\nu$, and is written as:
\begin{equation}
 {T}_{\nu} = \eket{c_\nu}\ebra{\bar{c}_\nu} + h.c.;
\end{equation}
in the simplest case, $ t{T_\nu}$ involves the minimum non-trivial dimer dynamics that preserve the constraint. Similarly, the potential energy term provides an energy cost for states with these local configurations:
\begin{equation}
 {V}_{\nu} = \eket{c_\nu}\ebra{c_\nu} +  \eket{\bar{c}_\nu}\ebra{\bar{c}_\nu}.
\end{equation}
Therefore we may define a Hamiltonian for a GQDM using \Eqref{eq:HGQDM} and defining a set of dimer resonances $\{(c_\nu,\bar{c}_\nu)\}$ and their corresponding strengths $\{(t_\nu,v_\nu)\}$:
\begin{equation}
\HGQDM \biggl( \bigl\{\left(c_\nu,\bar{c}_\nu\right)\},\{\left(t_\nu,v_\nu\right)\bigr\}\biggr).
\end{equation}
 In general the $[T_\nu, T_{\nu'}] \neq 0 $ and $[T_\nu,V_{\nu'}]\neq0$ when $\nu$ and $\nu'$ share links, and thus the ground state of $H_{\rm{GQDM}}$ is not explicitly known.

However, at the special points in the $\{(t_\nu,v_\nu)\}$ parameter space, known as the Rokhsar-Kivelson (RK) points, when \Eqref{eq:HGQDM} can be written as the sum of projection operators
\begin{equation}
H_{\rm{RK}} = \sum_\nu  {h}_\nu \label{eq:HRK},
\end{equation}
where all $h_\nu$ are projectors with eigenvalues $\{0,1\}$, the exact ground state is known to be the zero energy state of \Eqref{eq:HRK} that is simultaneously annihilated by all ${h}_\nu$. Such a zero energy state, known as the RK state, may be formed from an equal superposition of all dimerizations that are connected by $\{T_\nu\}$:
\begin{equation}
\eket{\Phi_{\RK}^{\Omega}} = \sum_{C \in \Omega} \eket{C}
\end{equation} 
where $\Omega$ is a subset of the Hilbert space connected by $\{T_\nu\}$. Thus $\{ \ket{\Phi_{\RK}^\Omega} \}$ are the exact, degenerate ground states of $H_{\rm{GQDM}}$ with a degeneracy labelled by $\Omega$ at the RK point.

@@

If a defect is introduced which violates the local constraint at a vertex, this can be viewed as an ``electric charge'' living on the vertex, within the context of the original GQDM. One may determine if such electric charges are confined by introducing a pair of such defects and computing their spatial correlation function; if the free energy of defects at large distances remains finite, electric charges are deconfined in the GQDM ground state.

The gauge symmetry allows for the construction of a gauge invariant magnetic string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell \label{eq:phi}
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv  \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) \tilde{n}_\ell \label{eq:phiBackground}
\end{equation}
where $n_\ell$ is the number operator for a static background dimer configuration which is chosen to fix the gauge. With this gauge fixing, the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma- \tilde{\Phi}_\Gamma$. An alternative physical picture is that there are static charges of charge $+c_0$ ($-c_0$) sitting at the vertices of the $A$($B$) sub-lattice, which are the source of the non-vanishing $\Phi_\Gamma$ flux through a closed loop.

We now may define the string operator
\begin{equation}
S_\gamma \left( \alpha \right) \equiv e^{i \alpha \left( \Phi_\gamma - \tilde{\Phi}_\gamma \right)} \label{eq:Sgamma}
\end{equation}
where $\alpha$ such that $S_\gamma$ commutes with $\HGQDM$ everywhere except at the ends of $\gamma$. Therefore, for closed loops $\Gamma$, $[ S_\Gamma,\HGQDM]=0$. For open strings, $S_\gamma$ does not commute with the Hamiltonian and can be viewed as creating a pair of ``magnetic charges'' at the ends of the string: correspondingly $S_\gamma \ket{\Phi_{\RK}}$ can used as a trial wavefunction for an excited state with a pair of magnetic vortices. If a state has topological order with deconfined electric charges, these magnetic vortices should be gapped excitations, and correspondingly the magnetic string operator should decay exponentially~\cite{Read1989a,Senthil2000,Senthil2001e}. The condensation of magnetic vortices can drive a phase transition from a topologically ordered phase to a conventionally order phase~\cite{Jalabert1991,Ralko2007,Huh2011}. Thus, the magnetic string correlation can be used as test for deconfinement of electric charges.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tpb]
    \centering
    \includegraphics[width=1.0\columnwidth]{mag_loops_on_orntd_lat.pdf}
    \caption{The A(B) sublattices of the square lattice are colored red (blue) and the link orientations are chosen to point from the A to the B sublattice. The narrow dark dimers are a static background dimerization used to fix the gauge, in this case chosen to be a columnar pattern; the wider light gray dimers are the physical degrees of freedom. $W_Y$ is a magnetic string operator with a non-trivial winding around one axis of the torus that measured the topological sector; $S_\gamma$ and $S_\Gamma$ are examples of open and closed string operators, respectively. Note the monomer defect which can be detected by $S_\Gamma$. \note{Owen: The sublattice coloring is inconsistent}}
    \label{fig:orientedSL}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The two unique fixed-number constraints on the square lattice are lead to QDM in the odd sector ($n_v = 1$) and the quantum full-packed loop model (QFPLM) in the even sector ($n_v = 2$). For configurations that satisfy these a number constraints, $Q_v$ strictly vanishes, and thus all states in the Hilbert space are invariant under $G_v$ for $0 \leq \alpha_v \leq 2 \pi$ in \Eqref{eq:Gv}; this represents a local $U(1)$ gauge symmetry~\cite{Moessner2001}.

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes $\bplqt$ of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have chosen $t_\bplqt = t$, $v_\bplqt =v, \forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0
\end{equation}
and therefore we can define two conserved topological winding numbers $\Phi_{X}$ \& $\Phi_{Y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On an lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_X$ and $\Phi_Y$ are integers in the range
\begin{equation}
-L_{x}/2 \leq  \Phi_{X} \leq L_{x}/2,
\end{equation}
and similarly for $\Phi_Y$, thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

At the Rokhsar-Kivelson point $t=v$, we may write $\HQDM$ in the form of \Eqref{eq:HRK} by defining projection operators  $ {h}_\nu \equiv  {P}^\dagger_{\nu} {P}_{\nu}$, with
\begin{equation}
 {P}^\dagger_\nu \equiv \biggl( \eket{ c_\nu} - \eket{\bar{c}_\nu} \biggr).
\end{equation} 
Since the plaquette flip term in \Eqref{eq:HQDM} is ergodic in each of topological sector $(\Phi_x,\Phi_y)$ on the square lattice~\cite{????}, each topological sector corresponds to a unique RK ground state where $\Omega$ is labeled by $(\Phi_x,\Phi_y)$. Thus the  RK points of the QDM and QFPLM have an \emph{extensively} degenerate ground states.

Defects in these models that violate the local dimer number constraint and thus carry non-zero U(1) electric charge. \tabref{tab:defects} lists all the possible defects and their associated U(1) charges in QDM and QFPLM. When a defect lives on sublattice A, it has the opposite charge that the same defect on sublattice B has. Note that in the QDM, the maximum charge is $\pm 3$, and there are six distinct possible non-trivial electric charges of defects.

The QDM and QFPLM on the square lattice have isolated liquid ground states at the RK point surrounded by symmetry broken phases~\cite{Rokhsar1988,Leung1996,Syljuasen2005,Syljuasen2006,Syljuasen2004,Moessner2008}. The RK point of these models have been shown to have power-law decaying dimer-dimer and monomer-monomer correlation functions and gapless excitations~\cite{Fisher1961,Kasteleyn1961,Fisher1966a,Fisher1963,Kasteleyn1963a,Sutherland1988b,Sutherland1988a,Kohmoto1988a,Liang1988,Sutherland1988,Kohmoto1988,Levitov1990,Leung1996,Henley1997,Henley1997,Henley2004a,Pollmann2006}. The algebraic decay of monomer correlations is indicative of the logarithmic confinement of electric charges in these models. Thus, such fixed number constrained models on bipartite lattices are expected to have an isolated unstable gapless critical quantum liquid points with logarithmically confined defect and an extensive topological degeneracy. These features of fixed number constrained models are consistent with the effective U(1) gauge theory description of the RK point~\cite{Fradkin2004,Fradkin1990,Moessner2001}; in 2+1D pure U(1) gauge theories are always confining and thus the lack of an extended deconfined topologically ordered phase near the RK point~\cite{PolyakovBook}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \begin{tabular}{| l | c | c | c | c | c |}
  	\hline
    %type & $n_v$ & $Q_v^{\rm{QDM}}$ & $Q_v^{\rm{FPLM}}$  & $Q_v^{Z_3}$ & \\
    type & $n_v$& configuration & $Q_v$ & $Q_v^{Z_2}$  & $Q_v^{\rm{Z_3}}$  \\
	\hline
    monomer & $0$& $\svrtmon$ & -1 & 1 &  -1   \\
	\hline
    dimer  & $1$ & $\svrtdmr$& 0 & 0 &   0  \\
	\hline
    trimer  & $2$ & $\svrttrmr$ $\svrttrmrtwo$& 1 & 1 &   1  \\
	\hline
    tetramer & $3$& $\svrtquat$  & 2 & 0 &   -1  \\
	\hline
    pentamer & $4$& $\svrtpent$ & 3 & 1 &  0   \\
	\hline
  \end{tabular}
  \caption{Local dimer configurations where $n_v$ dimers touch a vertex; only one representative of symmetry related configurations is shown. $Q_v$, $Q_v^{Z_2}$, and the $Q_v^{Z_3}$ are the associated $U(1)$, $Z_2$ and $Z_3$ charges on the A sub-lattice in the odd sector theories (QDM, odd IGT, QDPM). Defects on the B sub-lattice have opposite charge, with the exception of $Q_v^{Z_2}$, as there is only one type of non-trivial $Z_2$ charge.}
  \label{tab:defects}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We now consider relaxing the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) with constraints $n_v \in \{0,2,4\}$ and $n_v \in \{1,3\}$, respectively. In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{IGT}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{IGT}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

The increased dynamics of $\HIGT$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HIGT$. Consequently, to define the string and loop operators that commute with $\HIGT$ (for closed strings) we fix $\alpha = \pi$ in \Eqref{eq:Sgamma}:
\begin{equation}
S_\gamma^{\rm{IGT}} \equiv S_\gamma \left( \pi \right), \quad W_{X,Y}^{\rm{IGT}} \equiv W_{X,Y} \left( \pi \right).
\end{equation}
Thus the $W_{X,Y}^{\rm{IGT}}$ have two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as~$\Omega: (\pm 1,\pm1)$; consequently the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

Defects violating the dimer parity constraint in the IGT's have $Q_v^{Z_2} = 1$, as listed in \tabref{tab:defects}; we identify these defects as $Z_2$ electric charges on the corresponding vertices. Note that there is only one non-trivial charge in $Z_2$ gauge theory, thus there is no distinction between the charges of defects living on the two sublattices.

The ground states of $\HIGT$ are fourfold degenerate on the torus, and have exponentially decaying dimer-dimer and monomer-monomer correlations and a finite energy gap; such features demonstrate the deconfined topologically ordered nature of the ground state of the IGTs.
